Windowed least-squares model reduction for dynamical systems

### 2019-10-07 section4/spatial_numerical.tex
latest

Direct approaches solve~\eqref{eq:tclsrm} by
``transcribing" the infinite dimensional optimization problem into a finite
dimensional one by discretizing the state and objective function in time.
\KTC{Passive voice:} \EP{I think this reads clearly} The minimization problem is then reformulated as a nonlinear programming, or
optimization, problem. A variety of direct solution approaches exist, 
including collocation approaches, orthogonal polynomial (spectral  method) 
approaches, and genetic algorithms.  
\KTC{is this a
spectral method, or the spectral method? clarify} \EP{I don't understand what you mean. What is ``the" spectral method}

\KTC{you can't devide (i.e., decompose) a window into instances. you can divide a window into
sub-windows. you can also simply define a set of time instances in the window.
no need to emphasize `strictly increasing', as any set of points in the window
will by necessity be ordered.}

In the present context, the most straightforward direct solution approach consists of the 
following steps: (1) numerically discretize the FOM ODE (and hence the \textit{integrand} of~\eqref{eq:obj_gen_slab}) and 
(2) select a numerical quadrature rule to evaluate the \textit{integral} in~\eqref{eq:obj_gen_slab}.
To achieve this agenda, first define a discrete temporal grid 
by defining
$\nstepsArg{n}+1$ strictly increasing time instances described by
$\{\timeWindowArg{n}{i}\}_{i=0}^{\nstepsArg{n}}$ with
%\KTC{use cdots, not
%hdots or ldots. Don't use $\defeq$ for the first and last times; these are
%just equalities, i.e., you happen to set them equal to those values, but they
%aren't identically the same:}
\begin{align}
&\timeWindowArg{n}{0} \le \ldots \le \timeWindowArg{n}{\nstepsArg{n}} , \label{eq:timegrid1} \;
\timeWindowArg{n}{0} = \timeStartArg{n},\;
\timeWindowArg{n}{\nstepsArg{n}} = \timeEndArg{n} . 
%\timeWindowArg{n}{0} = \timeStartArg{n}, \timeWindowArg{n}{\nstepsArg{n}} =
%\timeEndArg{n}
\end{align}
Figure~\ref{fig:slab_fig2} depicts this discretization.
We now outline the direct solution approach for linear-multistep schemes; we note that the formulation for
other time-stepping methods (e.g., Runge--Kutta) follows closely. 
\begin{figure} 
\begin{centering} 
\includegraphics[trim={0.0cm 5cm 0cm 3cm},clip,width=1.0\textwidth]{figs/time_grid_timesteps.pdf} 
	\caption{Partitioning of time windows into time-step instances. \KTC{Don't call them `time-step instances'.
	This woudl be an instance of a time step. These are `time instances',
	correct? If so, fix this terminology everywhere.}} 
\label{fig:slab_fig2} 
\end{centering} 
\end{figure}
%We now outline several commonly employed temporal discretization 
%schemes, specifically linear multistep and collocation (Runge--Kutta) methods, that leverage this time-grid to discretize the FOM ODE and 
%objective functional.

@@

\residArg{n,i} &: (\stateyDiscreteArgnt{n,i},\ldots,\stateyDiscreteArgnt{n,i-k(i)}) \mapsto  \frac{1}{\Delta t^{n,i}} \sum_{j=0}^{\lmsWidthArg{n}{i}} \alpha^{n,i}_j \stateyDiscreteArgnt{n,i-j} -  \sum_{j=0}^{\lmsWidthArg{n}{i}} \beta^{n,i}_j \velocity(\stateyDiscreteArgnt{n,i-j},\timeWindowArg{n}{i-j},\tau^{n,i-j}) \\
               &: \RR{\fomdim} \otimes \RR{k(i)+1} \rightarrow \RR{\fomdim}. 

@@

Over each window, \directMethodAcronym\ computes the generalized coordinates $\approxstateDiscreteArg{n,0},\ldots,\approxstateDiscreteArg{n,\nstepsArg{n}}$, for $n=1,\ldots,\nslabs$, as the solution to the minimization problems,

@@

\begin{cases} \spatialIC & n = 2,\ldots,\nslabs,\\
\genstateICOne & n=1. \end{cases} \end{align*}

@@

To describe the Gauss--Newton method, we first define a ``vectorization" function that takes a tuple of time-local solutions at each time-step instance and outputs 
a concatenated vector, 
\begin{align*}
 \unroll &\vcentcolon (\stateyDiscreteArg{1},\ldots,\stateyDiscreteArg{m} ) \mapsto \begin{bmatrix} [\stateyDiscreteArg{1}]^T & \ldots & [\stateyDiscreteArg{m}]^T \end{bmatrix}^T  \\
&\vcentcolon \RR{p} \otimes \RR{m} \rightarrow \RR{pm},
\end{align*}
for arbitrary $p$ and $m$. The stacked vector of generalized coordinates over the $n$th time window are then defined as, 
\begin{equation*}
\genstatecollocMatSlabArg{n} \defeq 
\unroll (\genstateDiscreteArg{n,1},\ldots,\genstateDiscreteArg{n,\nstepsArg{n}} ).
\end{equation*}
Now, define the weighted space--time residual over the entire window as
\begin{equation*}
\residLMSSlabArg{n} : \genstatecollocMatySlabArg{} \mapsto \begin{bmatrix}
 \sqrt{\quadWeightsLMSScalarArg{n}{1} } \stweightingMatOneArg{n,1} \residLMSArg{n,1}( \basisspaceArg{n} \genstateyDiscreteArgnt{1} + \stateInterceptArg{n},  \basisspaceArg{n} \genstateyDiscreteArgnt{0} + \stateInterceptArg{n},) \\
\vdots \\
 \sqrt{\quadWeightsLMSScalarArg{n}{\nstepsArg{n}} } \stweightingMatOneArg{n,\nstepsArg{n}} \residLMSArg{n,\nstepsArg{n}}( \basisspaceArg{n} \genstateyDiscreteArgnt{\nstepsArg{n}} + \stateInterceptArg{n}, \ldots,\basisspaceArg{n} \genstateyDiscreteArgnt{\nstepsArg{n} - k^n(\nstepsArg{n}) } + \stateInterceptArg{n}) \\
\end{bmatrix},
\end{equation*}
where $\genstatecollocMatySlabArg{} \defeq \unroll (\genstateyDiscreteArg{0},\ldots,\genstateyDiscreteArg{\nstepsArg{n}} ).$
It is worth noting that, by design,
\begin{equation*}
\objectiveArgLMS{n} \bigg( \basisspaceArg{n} \genstateDiscreteArg{n,0} + \stateInterceptArg{n},\ldots,\basisspaceArg{n} \genstateDiscreteArg{n,\nstepsArg{n}} + \stateInterceptArg{n} \ \bigg) 
\equiv
\bigg[\residLMSSlabArg{n}  (\genstatecollocMatSlabArg{n}) \bigg]^T \bigg[ \residLMSSlabArg{n}(\genstatecollocMatSlabArg{n}) \bigg].
\end{equation*} 
Using these definitions, Algorithm~\ref{alg:colloc_gn} presents the standard Gauss--Newton method. The algorithm consists of three fundamental steps: (1) compute the FOM O$\Delta$E residual given the current guess, (2) compute the time window residual-Jacobian and form the normal equations, and (3) solve the normal equations and update the state.

@@

\begin{remark}\label{remark:gaussnewton}\textit{(Acceleration of the Gauss--Newton Solve)}\\
The principle cost of a Gauss--Newton method is often the formation of the Jacobian matrix. A variety of techniques aimed at 
reducing this computational burden exist; Jacobian-free Newton-Krylov methods~\cite{jfnk}, Broyden's method~\cite{broyden} (as explored in Ref.~\cite{carlberg_thesis}, Appendix A), and frozen Jacobian approximations are several such examples. Further, the space--time formulation exposes an extra dimension for parallelization that can be exposed to accelerate the wall-clock time of the ROM. The investigation of these additional, potentially more efficient, solution algorithms will be a topic of future work. 
\end{remark}
%\subsubsection{Jacobian-Free Implementation}
%The dominant cost of the Gauss Newton algorithm is the computatation of the action of the Jacobian matrix on the trial basis. To alleviate this burden, the following Jacobian-free method is proposed:
\begin{algorithm}
\caption{\spatialAcronym\ trial subspace: algorithm for the direct solution technique with the Gauss--Newton method and a linear multistep method over the $n$th window}
\label{alg:colloc_gn}
\SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}
\Input{tolerance, $\text{tol}$; initial guess, $\genstateGuessDiscreteArg{n,1}{0},\ldots,\genstateGuessDiscreteArg{n,\nstepsArg{n}}{0}$; initial condition $\genstateDiscreteArg{n,0}$}
\Output{Solution to least squares problem, $\genstatecollocMatSlabArg{n}$} 
\textbf{Online Steps}: \\
$\text{converged} \leftarrow \text{false}$ \Comment{Set convergence checker} \\
$k \leftarrow 0$ \Comment{Set counter}\\
$\genstatecollocMatSlabArg{n}_k \leftarrow \unroll(\genstateDiscreteArg{n,0},\genstateGuessDiscreteArg{n,1}{0},\ldots,\genstateGuessDiscreteArg{n,\nstepsArg{n}}{0})$ \Comment{Assemble generalized coordinates over window} \\
\While{\text{converged} == \text{false}}
{
%\For{$i=1,\hdots,\nstepsArg{n}$}{
%  \For{$j=1,\hdots,\ncollocArg{n}{i}$}{
%Compute: $\approxstateArgnt{n,i}  =  \basisspace  \genstateDiscreteArg{n,i} + \stateIntercept$ 
%\Comment{Compute state} \\
%Compute: $\velocity(\basisspace \genstateDiscreteArg{n,i} + \stateIntercept ) $ \Comment{Compute velocity at each time-step instance}\\
%Compute: $\residLMSArg{n,i}(\basisspace \genstateDiscreteArg{n,i} + \stateIntercept, \ldots , \basisspace \genstateDiscreteArg{n,i - k^n(i)} + \stateIntercept) $  \Comment{Compute residual} \\
%}
$\mathbf{r} \leftarrow \residLMSSlabArg{n}(\genstatecollocMatSlabArg{n}_k)$ \Comment{Compute weighted residual over window} \\
$\mathbf{J} \leftarrow  
\frac{\partial \residLMSSlabArg{n}}{\partial \genstatecollocMatySlabArg{}}(\genstatecollocMatSlabArg{n}_k) 
$ \Comment{Compute weighted residual-Jacobian over window} \\
%Compute: $[\jacobianSlabArg{n}]^T \jacobianSlabArg{n}$ \Comment{Compute system matrix for the normal equations} \\
%Compute: $[\jacobianSlabArg{n}]^T \residLMSSlabArg{n}(\genstatecollocMatSlabArg{n}_k)$ \Comment{compute RHS for normal equations} \\
$\Delta \genstatecollocMatSlabArg{n} \solveAlgo [\mathbf{J}]^T \mathbf{J} \Delta \genstatecollocMatSlabArg{n}=  -\mathbf{J}^T\mathbf{r}$ \Comment{Solve the normal equations} \\
$\genstatecollocMatSlabArg{n}_{k+1} \leftarrow \genstatecollocMatSlabArg{n}_k + \Delta \genstatecollocMatSlabArg{n}$ \Comment{Update guess to the state} \\
\If{ $\norm{ \mathbf{J}^T\mathbf{r}  } \le \text{tol}$ }{
{\text{converged} $\leftarrow$ \text{true}}  \Comment{Check and set convergence based on gradient norm} \\
Return: $\genstatecollocMatSlabArg{n} = \genstatecollocMatSlabArg{n}_{k+1} $ \Comment{Return converged solution}\\
}
$k\leftarrow k+1$
}
\end{algorithm}

@@

\subsection{Spatial trial space: indirect solution approach}\label{sec:indirect}
In contrast to the direct approach,
indirect methods ``indirectly" solve the minimization
problem~\eqref{eq:tclsrm} by solving the Euler-Lagrange equations. The
system given by the Euler Lagrange equations~\eqref{eq:lspg_continuous}-\eqref{eq:lspg_adjoint} comprises a coupled two-point boundary value
problem. Similar to the direct solution approach, a variety of techniques have
been devised to solve two-point boundary value problems of this type. These
techniques include shooting methods, multiple shooting
methods~\cite{multiple_shooting}, and the forward-backward sweep
method~\cite{fbs} (FBSM).  This work explores using the FBSM.

@@

The FBSM is an iterative approach that can be used to
solve the two-point boundary value problem. The general process of the FBSM is
as follows: First, the system~\eqref{eq:lspg_continuous} is (numerically) solved forward in
time using an initial guess to the costate state to obtain an initial guess to
the generalized coordinates. Next, the adjoint equation~\eqref{eq:lspg_adjoint} is solved \textit{backwards} in time given the
approximation to the generalized coordinates. This gives a new estimate to the
costate, which is then used to again solve the forward problem to obtain
a new estimate of the generalized coordinates. This process is continued until
convergence. Algorithm~\ref{alg:st_iter} outlines the method. The algorithm
contains three parameters: the damping factor $\rho$, the growth factor
$\fbsmGrowth$, and the decay factor $\fbsmDecay$. The damping factor controls the rate at which the costate seen by~\eqref{eq:lspg_continuous} is updated. The
closer $\rho$ is to unity, the faster the resulting algorithm will converge.
For large window sizes, however, too high a value of $\rho$ can lead to an unstable iterative process. 
A proper value of $\rho$ can be
obtained with a line search. The line search presented in~\eqref{eq:lspg_continuous} adapts the damping factor
according to the objective. Convergence properties of the FBSM method are
presented in Ref.~\cite{McAsey2012ConvergenceOT}. It is shown that, for a small 
enough value of $\rho$, the algorithm will converge.

>>>>

Direct approaches solve~\eqref{eq:tclsrm} by
``transcribing" the infinite dimensional optimization problem into a finite
dimensional one by discretizing the state and objective function in time.
The minimization problem is then reformulated as a nonlinear programming, or
optimization, problem. A variety of direct solution approaches exist, 
including collocation approaches, orthogonal polynomial (spectral  method) 
approaches, and genetic algorithms.  
\KTC{is this a
spectral method, or the spectral method? clarify} \EP{I don't understand what you mean. What is ``the" spectral method}

In the present context, the most straightforward direct solution approach consists of the 
following steps: (1) numerically discretize the FOM ODE (and hence the \textit{integrand} of~\eqref{eq:obj_gen_slab}) and 
(2) select a numerical quadrature rule to evaluate the \textit{integral} in~\eqref{eq:obj_gen_slab}.
To achieve this agenda, first define a discrete temporal grid 
by introducing
$\nstepsArg{n}+1$ strictly increasing time instances described by
$\{\timeWindowArg{n}{i}\}_{i=0}^{\nstepsArg{n}}$ with
\EP{These should be strictly increasing}
%\KTC{use cdots, not
%hdots or ldots. Don't use $\defeq$ for the first and last times; these are
%just equalities, i.e., you happen to set them equal to those values, but they
%aren't identically the same:}
\begin{align}
&\timeWindowArg{n}{0} \le \ldots \le \timeWindowArg{n}{\nstepsArg{n}} , \label{eq:timegrid1} \;
\timeWindowArg{n}{0} = \timeStartArg{n},\;
\timeWindowArg{n}{\nstepsArg{n}} = \timeEndArg{n} . 
%\timeWindowArg{n}{0} = \timeStartArg{n}, \timeWindowArg{n}{\nstepsArg{n}} =
%\timeEndArg{n}
\end{align}
Figure~\ref{fig:slab_fig2} depicts this discretization.
We now outline the direct solution approach for linear-multistep schemes; we note that the formulation for
other time-stepping methods (e.g., Runge--Kutta) follows closely. 
\begin{figure} 
\begin{centering} 
\includegraphics[trim={0.0cm 5cm 0cm 3cm},clip,width=1.0\textwidth]{figs/time_grid_timesteps.pdf} 
	\caption{Depiction of the $\nstepsArg{n}+1$ time instances over each window. In the figure, $\nstepsArg{n} = 3$ for all $n$.} 
\label{fig:slab_fig2} 
\end{centering} 
\end{figure}
%We now outline several commonly employed temporal discretization 
%schemes, specifically linear multistep and collocation (Runge--Kutta) methods, that leverage this time-grid to discretize the FOM ODE and 
%objective functional.

@@

\residArg{n,i} &: (\stateyDiscreteArgnt{i},\ldots,\stateyDiscreteArgnt{i-\lmsWidthArg{n}{i}}) \mapsto  \frac{1}{\Delta t^{n,i}} \sum_{j=0}^{\lmsWidthArg{n}{i}} \alpha^{n,i}_j \stateyDiscreteArgnt{i-j} -  \sum_{j=0}^{\lmsWidthArg{n}{i}} \beta^{n,i}_j \velocity(\stateyDiscreteArgnt{i-j},\timeWindowArg{n}{i-j}) \\
               &: \RR{\fomdim} \otimes \RR{\lmsWidthArg{n}{i}+1} \rightarrow \RR{\fomdim}. 

@@

Over each window, \methodAcronym\ computes the generalized coordinates $\approxstateDiscreteArg{n,0},\ldots,\approxstateDiscreteArg{n,\nstepsArg{n}}$, for $n=1,\ldots,\nslabs$, as the solution to the minimization problems,

@@

\begin{cases} \basisspaceArg{n} \spatialIC + \stateInterceptArg{n}& n = 2,\ldots,\nslabs,\\
\basisspaceArg{1} \genstateICOne  + \stateInterceptArg{1}& n=1. \end{cases} \end{align*}

@@

To describe the Gauss--Newton method, we first define a ``vectorization" function that takes a tuple of time-local solutions at each time-step instance and outputs 
a concatenated vector, 
\begin{align*}
 \unroll &\vcentcolon (\stateyDiscreteArg{1},\ldots,\stateyDiscreteArg{m} ) \mapsto \begin{bmatrix} [\stateyDiscreteArg{1}]^T & \ldots & [\stateyDiscreteArg{m}]^T \end{bmatrix}^T  \\
&\vcentcolon \RR{p} \otimes \RR{m} \rightarrow \RR{pm},
\end{align*}
for arbitrary $p$ and $m$. The stacked vector of generalized coordinates over the $n$th time window are then defined as, 
\begin{equation*}
\genstatecollocMatSlabArg{n} \defeq 
\unroll (\genstateDiscreteArg{n,1},\ldots,\genstateDiscreteArg{n,\nstepsArg{n}} ).
\end{equation*}
Now, define the weighted space--time residual over the entire window as
\begin{equation*}
\residLMSSlabArg{n} : (\genstatecollocMatySlabArg{};\genstateDiscreteArg{n,0}) \mapsto \begin{bmatrix}
 \sqrt{\quadWeightsLMSScalarArg{n}{1} } \stweightingMatOneArg{n,1} \residLMSArg{n,1}( \basisspaceArg{n} \genstateyDiscreteArgnt{1} + \stateInterceptArg{n},  \basisspaceArg{n} \genstateDiscreteArg{n,0} + \stateInterceptArg{n},) \\
\vdots \\
 \sqrt{\quadWeightsLMSScalarArg{n}{\nstepsArg{n}} } \stweightingMatOneArg{n,\nstepsArg{n}} \residLMSArg{n,\nstepsArg{n}}( \basisspaceArg{n} \genstateyDiscreteArgnt{\nstepsArg{n}} + \stateInterceptArg{n}, \ldots,\basisspaceArg{n} \genstateyDiscreteArgnt{\nstepsArg{n} - k^n(\nstepsArg{n}) } + \stateInterceptArg{n}) \\
\end{bmatrix},
\end{equation*}
where $\genstatecollocMatySlabArg{} \defeq \unroll (\genstateyDiscreteArg{1},\ldots,\genstateyDiscreteArg{\nstepsArg{n}} ).$
It is worth noting that, by design,
\begin{equation*}
\objectiveArgLMS{n} \bigg( \basisspaceArg{n} \genstateDiscreteArg{n,0} + \stateInterceptArg{n},\ldots,\basisspaceArg{n} \genstateDiscreteArg{n,\nstepsArg{n}} + \stateInterceptArg{n} \ \bigg) 
\equiv
\bigg[\residLMSSlabArg{n}  (\genstatecollocMatSlabArg{n};\genstateDiscreteArg{n,0}) \bigg]^T \bigg[ \residLMSSlabArg{n}(\genstatecollocMatSlabArg{n};\genstateDiscreteArg{n,0}) \bigg].
\end{equation*} 
Using these definitions, Algorithm~\ref{alg:colloc_gn} presents the standard Gauss--Newton method. The algorithm consists of three fundamental steps: (1) compute the FOM O$\Delta$E residual given the current guess, (2) compute the time window residual-Jacobian and form the normal equations, and (3) solve the normal equations and update the state.

@@

\begin{remark}\label{remark:gaussnewton}\textit{(Acceleration of the Gauss--Newton Solve)}\\
The principle cost of a Gauss--Newton method is often the formation of the Jacobian matrix. A variety of techniques aimed at 
reducing this computational burden exist; Jacobian-free Newton-Krylov methods~\cite{jfnk}, Broyden's method~\cite{broyden} (as explored in Ref.~\cite{carlberg_thesis}, Appendix A), and frozen Jacobian approximations are several such examples. Further, the space--time formulation exposes an extra dimension for parallelization that can be exposed to accelerate the wall-clock time of the ROM. The investigation of these additional, potentially more efficient, solution algorithms will be a topic of future work. 
\end{remark}
%\subsubsection{Jacobian-Free Implementation}
%The dominant cost of the Gauss Newton algorithm is the computatation of the action of the Jacobian matrix on the trial basis. To alleviate this burden, the following Jacobian-free method is proposed:
\begin{algorithm}
\caption{\spatialAcronym\ trial subspace: algorithm for the direct solution technique with the Gauss--Newton method and a linear multistep method over the $n$th window}
\label{alg:colloc_gn}
\SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}
\Input{tolerance, $\epsilon$; initial guess, $\genstateGuessDiscreteArg{n,1}{0},\ldots,\genstateGuessDiscreteArg{n,\nstepsArg{n}}{0}$; initial condition $\genstateDiscreteArg{n,0}$}
\Output{Solution to least squares problem, $\genstatecollocMatSlabArg{n}$} 
\textbf{Online Steps}: \\
$\text{converged} \leftarrow \text{false}$ \Comment{Set convergence checker} \\
$k \leftarrow 0$ \Comment{Set counter}\\
$\genstatecollocMatSlabArg{n}_k \leftarrow \unroll(\genstateGuessDiscreteArg{n,1}{0},\ldots,\genstateGuessDiscreteArg{n,\nstepsArg{n}}{0})$ \Comment{Assemble generalized coordinates over window} \\
\While{\text{converged} == \text{false}}
{
%\For{$i=1,\hdots,\nstepsArg{n}$}{
%  \For{$j=1,\hdots,\ncollocArg{n}{i}$}{
%Compute: $\approxstateArgnt{n,i}  =  \basisspace  \genstateDiscreteArg{n,i} + \stateIntercept$ 
%\Comment{Compute state} \\
%Compute: $\velocity(\basisspace \genstateDiscreteArg{n,i} + \stateIntercept ) $ \Comment{Compute velocity at each time-step instance}\\
%Compute: $\residLMSArg{n,i}(\basisspace \genstateDiscreteArg{n,i} + \stateIntercept, \ldots , \basisspace \genstateDiscreteArg{n,i - k^n(i)} + \stateIntercept) $  \Comment{Compute residual} \\
%}
$\mathbf{r} \leftarrow \residLMSSlabArg{n}(\genstatecollocMatSlabArg{n}_k;\genstateDiscreteArg{n,0})$ \Comment{Compute weighted residual over window} \\
$\mathbf{J} \leftarrow  
\frac{\partial \residLMSSlabArg{n}}{\partial \genstatecollocMatySlabArg{}}(\genstatecollocMatSlabArg{n}_k) 
$ \Comment{Compute weighted residual-Jacobian over window} \\
%Compute: $[\jacobianSlabArg{n}]^T \jacobianSlabArg{n}$ \Comment{Compute system matrix for the normal equations} \\
%Compute: $[\jacobianSlabArg{n}]^T \residLMSSlabArg{n}(\genstatecollocMatSlabArg{n}_k)$ \Comment{compute RHS for normal equations} \\
$\Delta \genstatecollocMatSlabArg{n} \solveAlgo [\mathbf{J}]^T \mathbf{J} \Delta \genstatecollocMatSlabArg{n}=  -\mathbf{J}^T\mathbf{r}$ \Comment{Solve the normal equations} \\
$\genstatecollocMatSlabArg{n}_{k+1} \leftarrow \genstatecollocMatSlabArg{n}_k + \Delta \genstatecollocMatSlabArg{n}$ \Comment{Update guess to the state} \\
\If{ $\norm{ \mathbf{J}^T\mathbf{r}  } \le \epsilon$ }{
{\text{converged} $\leftarrow$ \text{true}}  \Comment{Check and set convergence based on gradient norm} \\
Return: $\genstatecollocMatSlabArg{n} = \genstatecollocMatSlabArg{n}_{k+1} $ \Comment{Return converged solution}\\
}
$k\leftarrow k+1$
}
\end{algorithm}

@@

\subsection{Spatial trial space: indirect solution approach}\label{sec:indirect}
In contrast to the direct approach,
indirect methods ``indirectly" solve the minimization
problem~\eqref{eq:tclsrm} by solving the Euler--Lagrange equations. The
system given by the Euler--Lagrange equations~\eqref{eq:lspg_continuous}--\eqref{eq:lspg_adjoint} comprises a coupled two-point boundary value
problem. Similar to the direct solution approach, a variety of techniques have
been devised to solve two-point boundary value problems of this type. These
techniques include shooting methods, multiple shooting
methods~\cite{multiple_shooting}, and the forward-backward sweep
method~\cite{fbs} (FBSM).  This work explores using the FBSM.

@@

The FBSM is an iterative approach that can be used to
solve the two-point boundary value problem. The general process of the FBSM is
as follows: First, the system~\eqref{eq:lspg_continuous} is (numerically) solved forward in
time using an initial guess to the costate state to obtain an initial guess to
the generalized coordinates. Next, the adjoint equation~\eqref{eq:lspg_adjoint} is solved \textit{backwards} in time given the
approximation to the generalized coordinates. This gives a new estimate to the
costate, which is then used to again solve the forward problem to obtain
a new estimate of the generalized coordinates. This process is continued until
convergence. Algorithm~\ref{alg:st_iter} outlines the method. The algorithm
contains three parameters: the damping factor $\rho$, the growth factor
$\fbsmGrowth$, and the decay factor $\fbsmDecay$. The damping factor controls the rate at which the costate seen by~\eqref{eq:lspg_continuous} is updated. The
closer $\rho$ is to unity, the faster the resulting algorithm will converge.
For large window sizes, however, too high a value of $\rho$ can lead to an unstable iterative process. 
A proper value of $\rho$ can be
obtained with a line search. The line search presented in Algorithm~\ref{alg:st_iter} adapts the damping factor
according to the objective. Convergence properties of the FBSM method are
presented in Ref.~\cite{McAsey2012ConvergenceOT}. It is shown that, for a small 
enough value of $\rho$, the algorithm will converge.

\begin{algorithm} \caption{\spatialAcronym\ trial subspace: algorithm for the FBSM over the $n$th window.} \label{alg:st_iter} 
\SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}
\Input{tolerance, $tol$; damping factor, $\rho$; growth factor, $\fbsmGrowth$; decay
factor, $\fbsmDecay$} 
\Output{Stationary point, $\genstateArgnt{n}$ }
\textbf{Online Steps:}\\ 
$\genstate^n_0 \leftarrow \bz$ \Comment{Set initial guess for state} \\
$\controllerArgnt{n} \leftarrow \boldsymbol 0$ \Comment{Set initial
guess for costate}\\ 
$\genstateArgnt{n}_1 \solveAlgo \basisspaceTArg{n} \stweightingMatArg{n}
\basisspaceArg{n} \genstateDotArgnt{n}_1  -  \basisspaceTArg{n} \stweightingMatArg{n}
\velocity(\veloargsromntArg{1}) =  \controllerArgnt{n}$ 
% Eq.~\ref{eq:lspg_continuous} with $\adjoint^n(t) = \controllerArg{n}{t}$ to
% obtain $\genstate_0^{n}(t)$ 
\Comment{Solve~\eqref{eq:lspg_continuous} with initial guess to
controller}\\ 
%Set: $\genstate^n_1 = \genstate$ \Comment{Set state} \\
%$\objectiveArg{n}({\genstate_1^n(t)})$
%\Comment{Evaluate objective function}\\
$i \leftarrow 1$ \Comment{Set counter}\\
\While{$tol \le \int_{\timeStartArg{n}}^{\timeEndArg{n}} \norm{\genstate^n_{i}(t) - \genstate^n_{i-1}(t) }dt $}{
\small{
\begin{multline*}
\adjointArgnt{n} \underset{\text{solve}}{\leftarrow} 
\frac{d}{dt} \adjoint^n  + \basisspaceTArg{n} \bigg[\frac{\partial \velocity}{\partial \stateyDiscrete}(\basisspaceArg{n} \genstate^n_i + \stateIntercept,t) \bigg]^T \stweightingMatArg{n} \basisspaceArg{n} [\massArg{n}]^{-1} \adjoint^n = \\ -\bigg[\basisspaceTArg{n} \bigg[ \frac{\partial \velocity}{\partial \stateyDiscrete} ( \basisspaceArg{n} \genstate^n_i + \stateInterceptArg{n},t) \bigg]^T \stweightingMatArg{n} \bigg( \mathbf{I} -   \basisspace [\massArg{n}]^{-1} \basisspaceTArg{n} \bigg)  \stweightingMatArg{n} \bigg( \basisspaceArg{n} \dot{\genstate}_i^n   -   \velocity( \basisspaceArg{n} \genstate_i^n  +\stateInterceptArg{n},t) \bigg) \bigg] 
\end{multline*} }
\Comment{Solve~\eqref{eq:lspg_adjoint} to obtain guess to costate state} \\
%Solve Eq.~\ref{eq:lspg_adjoint} with $\genstate^n(t) = \genstate^n_i(t)$ to obtain $\adjoint^n(t)$ \\
$\controllerArgnt{n}  \leftarrow \rho \controllerArgnt{n} + (1 - \rho) \adjointArgnt{n}$ \Comment{Weighted update to costate}\\
$i \leftarrow i+1$ \Comment{Update Counter}\\
$\genstateArgnt{n}_i \solveAlgo \basisspaceTArg{n} \stweightingMatArg{n} \basisspaceArg{n} \genstateDotArgnt{n}_i   -  \basisspaceTArg{n} \stweightingMatArg{n} \velocity(\basisspaceArg{n} \genstate^n_i + \stateInterceptArg{n},t) =  \controllerArgnt{n} $
\Comment{Solve~\ref{eq:lspg_continuous} with updated guess to costate}\\
%$\objectiveArg{n}({\genstate_i^n(t)})$ \Comment{Evaluate objective function}\\
\uIf{ $\objectiveArg{n}({\genstate_i^n}) \le \objectiveArg{n}({\genstate_{i-1}^n})$}
{
$\rho \leftarrow \text{min}(\rho \fbsmGrowth,1)$ \Comment{Grow the damping factor}\\
}
\Else{
$\rho \leftarrow \frac{\rho }{ \fbsmDecay}$ \Comment{Shrink the damping factor}\\ 
$\genstate_i^{n} \leftarrow  \genstate_{i-1}^{n}$ \Comment{Reset state to value at previous iteration}
}
}
Return converged solution, $\genstateArgnt{n}= \genstateArgnt{n}_i$
\end{algorithm}
\subsubsection{Numerical considerations for the forward and backward problems}
The FBSM requires solving~\eqref{eq:lspg_continuous} and~\eqref{eq:lspg_adjoint}, both of which are defined at the time-continuous level. 
The numerical implementation of the FBSM requires two main ingredients: (1) temporal discretization schemes for the forward and backward problems and (2) 
an efficient method for computing action of the Jacobian transpose on a vector.

This work examines the use of linear multistep methods for the purpose of temporal discretization of the forward and 
backwards problems. As described in  
the direct solution approach, temporal discretization is achieved by introducing $\nstepsArg{n} + 1$ time-step instances~\eqref{eq:timegrid1} over each time window. Linear multistep methods then use this time-grid for temporal discretization of~\eqref{eq:lspg_continuous} and~\eqref{eq:lspg_adjoint}.

The second ingredient, namely an efficient method for computing the action of the Jacobian transpose on a vector, can be challenging. 
In the case one does not have access to the full-order model Jacobian (or it is too costly to compute), the evaluation of this term can be both challenging and expensive. Here, two methods are discussed that can be used to evaluate the action of the Jacobian transpose on a vector as seen in~\eqref{eq:lspg_adjoint}:
\begin{enumerate}
\item Jacobian-free Derivative Approximation: A non-intrusive way to evaluate the Jacobian transpose in~\eqref{eq:lspg_adjoint} is to recognize that all terms including the Jacobian transpose are left multiplied by the transpose of the trial basis. One can make the manipulation,
$$\basisspaceTArg{n} \bigg[\frac{\partial \velocity}{\partial \stateyDiscrete} (\veloargsromn)\bigg]^T = \bigg[  \frac{\partial \velocity}{\partial \stateyDiscrete} (\veloargsromn) \basisspaceArg{n} \bigg]^T.$$
This manipulation allows one to compute the action of the Jacobian on each column of $\basisspace$ by, e.g., a finite difference approximation,
$$\frac{\partial \velocity}{\partial \stateyDiscrete}(\veloargsromn) \basisvec_i^n \approx \frac{1}{\epsilon}\bigg( \velocity(\basisspaceArg{n}\genstateArg{n}{t} + \stateInterceptArg{n} + \epsilon \basisvec_i^n,t) - \velocity(\veloargsromn) \bigg).$$
The Jacobian transpose can be formed in $K+1$ evaluations of the velocity. For cases where the reduced-order model is low dimensional, this approach is feasible. For higher dimensional reduced-order models, this approach can be prohibitively expensive.
